---
title: "Optimal number of clusters"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load required packages
library(factoextra)
library(NbClust)
```

#template for determining and visualizing optimal number of clusters
https://rpkgs.datanovia.com/factoextra/reference/fviz_nbclust.html

#fviz_nbclust(x, FUNcluster = NULL, method = c("silhouette", "wss",
  "gap_stat"), diss = NULL, k.max = 10, nboot = 100,
  verbose = interactive(), barfill = "steelblue",
  barcolor = "steelblue", linecolor = "steelblue",
  print.summary = TRUE, ...)

#fviz_gap_stat(gap_stat, linecolor = "steelblue", maxSE = list(method =
  "firstSEmax", SE.factor = 1))
  
#once raw data is scaled (df.scaled), measures to find optimal number of clusters

```{r}
# Elbow method
fviz_nbclust(df.scaled, kmeans, method = "wss") + 
  # add line for better visualization
  labs(subtitle = "Elbow method") # add subtitle
```

```{r}
#The Silhouette method measures the quality of a clustering and determines 
#how well each point lies within its cluster.

# Silhouette method
fviz_nbclust(df.scaled, kmeans, method = "silhouette") +
  labs(subtitle = "Silhouette method")
```

```{r}
#Gap statistic method, bootstrapping method
set.seed(42)
fviz_nbclust(df, kmeans,
             nstart = ,
             method = "gap_stat",
             nboot = 500 # bootstrap samples, reduce it for lower computation time (but less precise results)
) +
  labs(subtitle = "Gap statistic method")

#alternative script
# Perform (enhanced) k-means clustering
km <- eclust(df, "kmeans", nstart = 25)
# Gap statistic plot
fviz_gap_stat(km$gap_stat)

#alternative script
# Perform k-means clustering using clusGap
library(cluster)
gap_stat <- clusGap(df, FUN = kmeans, 
                    nstart = 25,
                    K.max = 10, #maximum number of clusters
                    B = ) #insert number less than 500
 print(gap_stat, method = "firstmax")
fviz_gap_stat(gap_stat)
```

```{r}
# Print cluster result
km
```

```{r}
#The NbClust() function from the {NbClust} package, which provides 30 indices for choosing the best number of clusters.
nbclust_out <- NbClust(
  data = , #insert data name
  distance = "euclidean",
  min.nc = , # insert minimum number of clusters
  max.nc = , # insert maximum number of clusters
  method = "kmeans" # one of: "ward.D", "ward.D2", "single", "complete", "average", "mcquitty", "median", "centroid", "kmeans"
)
nbclust_out #outputs results from various indicies
fviz_nbclust(nbclust_out)
```
